Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2015-02-17 MS_revised.tex
ref to tables

The MEDUSA analysis on the MCC tree in combination with richness data
estimated 18 significant changes in the tempo of diversification in
Nymphalidae history (Fig.~\ref{fig1}; Table 1). The corrected AICc acceptance
threshold for adding subsequent piecewise birth-death processes to the
overall model was set to 7.8 units, as prescribed by MEDUSA. In all
MEDUSA analyses, the maximum number of inferred diversification shifts
in all trees was 26. The background net diversification rate for
Nymphalidae was estimated as \texttt{r = 0.092} lineages per Million of
years and the AICc score for the best fit model was \texttt{5090.5}
(Table 1). MEDUSA also estimated that the basic constant birth-death
model was not a better explanation for our data (AICc
\texttt{= 5449.3}).

@@

Some of the 18 changes in diversification correspond to rate increases
in very species-rich genera: \emph{Ypthima} (\texttt{r = 0.311}),
\emph{Charaxes} (\texttt{r = 0.291}), \emph{Callicore} +
\emph{Diaethria} (\texttt{r = 0.220}), \emph{Pedaliodes}
(\texttt{r = 0.196}) and \emph{Taenaris} (\texttt{r = 0.234}). We found
rate increases for other clades as well such as: Mycalesina
(\texttt{r = 0.191}), Oleriina + Ithomiina + Napeogenina + Dircennina +
Godyridina (\texttt{r = 0.187}), Satyrinae (\texttt{r = 0.116}),
Phyciodina in part (\texttt{r = 0.241}) and Satyrina
(\texttt{r = 0.221}), \emph{Coenonympha} (\texttt{r = 0.209}),
\emph{Caeruleuptychia} + \emph{Magneuptychia} (\texttt{r = 0.312}) and
\emph{Taenaris} (\texttt{r = 0.312}). We also found decreases in net
diversification rates for Limenitidinae + Heliconiinae
(\texttt{r = 0.0541}), part of Danaini (\texttt{r = 0.0423}),
Pseudergolinae (\texttt{r = 0.024}) and Coenonymphina
(\texttt{r = 0.065}) (Table 1).

@@

We found that the analysis by MultiMEDUSA on the 1000 trees estimated
lower median net diversification rates for the diversification shifts
found by MEDUSA on the MCC tree derived from the random sample of trees
(Table 2). Although the diversification pattern found by MEDUSA and
MultiMEDUSA was the same, the latter consistently estimated lower rates.
Furthermore, the shifts recovered with low net diversification rate on
the MCC were recovered with negative net diversification rate by
MultiMEDUSA. The background diversification and all shifts found by
MEDUSA on the 1000 trees are provided as an R object in S15 File.

@@

For the diversification shifts found in both the MCC tree and most of
the samples of 1000 trees (frequency more than 90\%; Table 2), the
MultiMEDUSA approach recovered different rates of diversification than
those found when the MCC tree alone was used.

There were four net diversification rate increases found in the trees from the
random sample (Table 2): (i) the
genus \emph{Ypthima} (\(r = 0.22\)); (ii) the genus \emph{Charaxes}
(\(r = 0.21\)); (iii) Ithomiini subtribes Oleriina + Ithomiina + Napeogenina +
Dircennina + Godyridina (\(r = 0.10\)); and (iv) the clade of 
\emph{Callicore} + \emph{Diaethria} (\(r = 0.135\)).

@@

The MEDUSA analyses, taking into account phylogenetic uncertainty,
estimated a net diversification rate increase in part of the clade
Ithomiini across more than 95\% of the trees. Our BiSSE analysis found a
positive effect of the character state ``feeding on Solanaceae'' on the
net net diversification rate on part of Ithomiini (Oleriina + Ithomiina
+ Napeogenina + Dircennina + Godyridina) (Fig.~\ref{fig4}). The estimated mean
net diversification rate for taxa that do not feed on Solanaceae was
\texttt{r = 0.11} while the net diversification rate for the Solanaceae
feeders was \texttt{r = 0.16} (see S1 Figure for a boxplot of speciation
and extinction values for the 95\% credibility intervals). The same
analysis considering \emph{Vanessa} and \emph{Hypanartia} as
non-Solanaceae feeders due to dubious records produced the same pattern
and net diversification rates (S2 Figure). Therefore, the rest of the
analyses were performed assuming these two genera as Solanaceae feeders.
We constrained the BiSSE likelihood model to force equal rates of
speciation for both character states in order to test whether the model
of different speciation rates is a significantly better explanation for
the data. A likelihood ratio test found that the model for increased net
diversification rate for nymphalids feeding on Solanaceae is a
significantly better explanation than this character state having no
effect on diversification (\(\chi^2 = 12.3; 1 df; p < 0.001\)) (Table 3;
character states available in S13 Dataset, code in S18 Supporting Information,
and mcmc run in S19 Supporting Information). We combined the post-burnin mcmc generations
from running BiSSE on 250 trees from the random sample of 1000 trees and
found the same pattern as the BiSSE analysis on the maximum clade
credibility tree (combined mcmc run in S20 Supporting Information; profiles
plot of speciation rates in S3 Figure; boxplot of 95\% credibility intervals in
S4 Figure). A BiSSE analysis to test whether the trait ``feeding on
Apocynaceae'' had any effect on increased net diversification rates
found similar speciation rates for lineages feeding on Apocynaceae and
other plants (S5 Figure).

@@

The MEDUSA method has been used to infer changes in net diversification
rates in a phylogenetic tree. Since its publication \cite{alfaro2009} the results
of using MEDUSA on a single tree, the maximum clade credibility tree,
have been used for generation of hypotheses and discussion
\cite{heikkila2012, litman2011, ryberg2012}. However, different diversification
shifts and different
rates of diversification are found for certain lineages when
phylogenetic uncertainty was taken into account by using MEDUSA on a
random sample of trees from the posterior distribution of a Bayesian
run. We found that some diversification shifts, estimated on the
Nymphalidae maximum clade credibility tree, were found with very low
probability in the random sample of 1000 trees from the posterior
distribution (Fig.~\ref{fig3}, Table 2). We also found that, even though the
analyses estimated the same diversification shifts on two or more trees,
the estimated net diversification rates could vary widely (Fig.~\ref{fig2}). For
example, in our Nymphalidae trees, we found that the shift for
\emph{Magneuptychia} and \emph{Caeruleuptychia} had a rate from
\texttt{r = 0.5234}, higher than the background net diversification
rate, to almost zero. This means that observed patterns and conclusions
can be completely contradictory depending on tree choice.

>>>>

The MEDUSA analysis on the MCC tree in combination with richness data
estimated 18 significant changes in the tempo of diversification in
Nymphalidae history (Fig.~\ref{fig1}; Table~\ref{table1}). The corrected AICc acceptance
threshold for adding subsequent piecewise birth-death processes to the
overall model was set to 7.8 units, as prescribed by MEDUSA. In all
MEDUSA analyses, the maximum number of inferred diversification shifts
in all trees was 26. The background net diversification rate for
Nymphalidae was estimated as \texttt{r = 0.092} lineages per Million of
years and the AICc score for the best fit model was \texttt{5090.5}
(Table~\ref{table1}). MEDUSA also estimated that the basic constant birth-death
model was not a better explanation for our data (AICc
\texttt{= 5449.3}).

@@

Some of the 18 changes in diversification correspond to rate increases
in very species-rich genera: \emph{Ypthima} (\texttt{r = 0.311}),
\emph{Charaxes} (\texttt{r = 0.291}), \emph{Callicore} +
\emph{Diaethria} (\texttt{r = 0.220}), \emph{Pedaliodes}
(\texttt{r = 0.196}) and \emph{Taenaris} (\texttt{r = 0.234}). We found
rate increases for other clades as well such as: Mycalesina
(\texttt{r = 0.191}), Oleriina + Ithomiina + Napeogenina + Dircennina +
Godyridina (\texttt{r = 0.187}), Satyrinae (\texttt{r = 0.116}),
Phyciodina in part (\texttt{r = 0.241}) and Satyrina
(\texttt{r = 0.221}), \emph{Coenonympha} (\texttt{r = 0.209}),
\emph{Caeruleuptychia} + \emph{Magneuptychia} (\texttt{r = 0.312}) and
\emph{Taenaris} (\texttt{r = 0.312}). We also found decreases in net
diversification rates for Limenitidinae + Heliconiinae
(\texttt{r = 0.0541}), part of Danaini (\texttt{r = 0.0423}),
Pseudergolinae (\texttt{r = 0.024}) and Coenonymphina
(\texttt{r = 0.065}) (Table~\ref{table1}).

@@

We found that the analysis by MultiMEDUSA on the 1000 trees estimated
lower median net diversification rates for the diversification shifts
found by MEDUSA on the MCC tree derived from the random sample of trees
(Table~\ref{table2}). Although the diversification pattern found by MEDUSA and
MultiMEDUSA was the same, the latter consistently estimated lower rates.
Furthermore, the shifts recovered with low net diversification rate on
the MCC were recovered with negative net diversification rate by
MultiMEDUSA. The background diversification and all shifts found by
MEDUSA on the 1000 trees are provided as an R object in S15 File.

@@

For the diversification shifts found in both the MCC tree and most of
the samples of 1000 trees (frequency more than 90\%; Table~\ref{table2}), the
MultiMEDUSA approach recovered different rates of diversification than
those found when the MCC tree alone was used.

There were four net diversification rate increases found in the trees from the
random sample (Table~\ref{table2}): (i) the
genus \emph{Ypthima} (\(r = 0.22\)); (ii) the genus \emph{Charaxes}
(\(r = 0.21\)); (iii) Ithomiini subtribes Oleriina + Ithomiina + Napeogenina +
Dircennina + Godyridina (\(r = 0.10\)); and (iv) the clade of 
\emph{Callicore} + \emph{Diaethria} (\(r = 0.135\)).

@@

The MEDUSA analyses, taking into account phylogenetic uncertainty,
estimated a net diversification rate increase in part of the clade
Ithomiini across more than 95\% of the trees. Our BiSSE analysis found a
positive effect of the character state ``feeding on Solanaceae'' on the
net net diversification rate on part of Ithomiini (Oleriina + Ithomiina
+ Napeogenina + Dircennina + Godyridina) (Fig.~\ref{fig4}). The estimated mean
net diversification rate for taxa that do not feed on Solanaceae was
\texttt{r = 0.11} while the net diversification rate for the Solanaceae
feeders was \texttt{r = 0.16} (see S1 Figure for a boxplot of speciation
and extinction values for the 95\% credibility intervals). The same
analysis considering \emph{Vanessa} and \emph{Hypanartia} as
non-Solanaceae feeders due to dubious records produced the same pattern
and net diversification rates (S2 Figure). Therefore, the rest of the
analyses were performed assuming these two genera as Solanaceae feeders.
We constrained the BiSSE likelihood model to force equal rates of
speciation for both character states in order to test whether the model
of different speciation rates is a significantly better explanation for
the data. A likelihood ratio test found that the model for increased net
diversification rate for nymphalids feeding on Solanaceae is a
significantly better explanation than this character state having no
effect on diversification (\(\chi^2 = 12.3; 1 df; p < 0.001\))
(Table~\ref{table3};
character states available in S13 Dataset, code in S18 Supporting Information,
and mcmc run in S19 Supporting Information). We combined the post-burnin mcmc generations
from running BiSSE on 250 trees from the random sample of 1000 trees and
found the same pattern as the BiSSE analysis on the maximum clade
credibility tree (combined mcmc run in S20 Supporting Information; profiles
plot of speciation rates in S3 Figure; boxplot of 95\% credibility intervals in
S4 Figure). A BiSSE analysis to test whether the trait ``feeding on
Apocynaceae'' had any effect on increased net diversification rates
found similar speciation rates for lineages feeding on Apocynaceae and
other plants (S5 Figure).

@@

The MEDUSA method has been used to infer changes in net diversification
rates in a phylogenetic tree. Since its publication \cite{alfaro2009} the results
of using MEDUSA on a single tree, the maximum clade credibility tree,
have been used for generation of hypotheses and discussion
\cite{heikkila2012, litman2011, ryberg2012}. However, different diversification
shifts and different
rates of diversification are found for certain lineages when
phylogenetic uncertainty was taken into account by using MEDUSA on a
random sample of trees from the posterior distribution of a Bayesian
run. We found that some diversification shifts, estimated on the
Nymphalidae maximum clade credibility tree, were found with very low
probability in the random sample of 1000 trees from the posterior
distribution (Fig.~\ref{fig3}, Table~\ref{table2}). We also found that, even though the
analyses estimated the same diversification shifts on two or more trees,
the estimated net diversification rates could vary widely (Fig.~\ref{fig2}). For
example, in our Nymphalidae trees, we found that the shift for
\emph{Magneuptychia} and \emph{Caeruleuptychia} had a rate from
\texttt{r = 0.5234}, higher than the background net diversification
rate, to almost zero. This means that observed patterns and conclusions
can be completely contradictory depending on tree choice.
